Generation of Variation and a Modified Mean Fitness Principle: \\ Necessity is the Mother of Genetic Invention

### 2018-02-11 ms.tex
MWF editing

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form 
%(introduced to describe conditional dispersal in \citet{Altenberg2012a})
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
$C$ is a positive diagonal  matrix with entries $C_k$, where $C_k<1$ for at least one $k$;
$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity, \citealp[see][Appendix 3]{Otto2007});
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pg.~203--206]{Altenberg1984}.

In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
This "absorption" ensures $M$ is column-stochastic. 
In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
The probability for maturation to other phenotypes is $0$.

@@

We first focus on the stable population mean fitness.
We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.  We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle involves a decrease in the mean fitness with increasing rates.

@@

\begin{result}[Modified mean fitness principle]\label{result:mfp}
Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding positive left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
Then,
\begin{equation}\label{eq:mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $\hat u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $\hat u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{eq:mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, since $\hat u_k, \hat v_k, C_k > 0$, we have
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

\begin{corollary}
The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
\begin{multline}\label{eq:mfp_K}
%\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
%\end{aligned}
\end{multline}
and
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
\end{multline}
\end{corollary}

@@

\begin{corollary}[Reproductive value principle]\label{cor:rvp}
In the notation of Result~\ref{result:mfp},
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

\begin{result}[Evolution of increased genetic variation]
Let $\lambda_1$ be the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $b$ and $B$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

Now, applying Result~\ref{result:mfp} (Eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
\end{equation}
Thus
\begin{equation}
\begin{aligned}
sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
sign(\hat{\bar w} - w_k),
\end{aligned}
\end{equation}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated. Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural selection.
These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness. 
This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be irreducible. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
We refer to this as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers that uniformly affect all individuals, the mean fitness here can increase as rates increase.

@@

Since $D$ is a positive diagonal matrix, and by assumption $C_k < 1$ for at least one $k$, $E$ has at least one positive diagonal element, making it aperiodic.  Further more, since $D$ is a positive diagonal matrix and $M$ is irreducible, $E$ is also irreducible.  Hence $E$ is primitive.

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form 
%(introduced to describe conditional dispersal in \citet{Altenberg2012a})
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
\citep{Kirkland2006,Altenberg2012b} where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
$C$ is a positive diagonal  matrix with entries $C_k$, where $C_k<1$ for at least one $k$;
$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity, \citealp[see][Appendix 3]{Otto2007});
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pp.~203--206]{Altenberg1984}.

In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any ``spillover'' of phenotypes below $1$ or above $n$ is ``absorbed'' by those boundaries.
This ``absorption'' ensures $M$ is column-stochastic. 
In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
The probability for maturation to other phenotypes is $0$.

@@

We first focus on the stable population mean fitness.
We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.  We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle~\citep{Karlin1972}  involves a decrease in the mean fitness with increasing rates.

@@

\begin{result}[Modified mean fitness principle]\label{result:mfp}
Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding positive left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
Then,
\begin{equation}\label{eq:mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:mfp_sign}
{\rm sign}\frac{\partial \hat{\bar w}}{\partial C_k} = 
{\rm sign}(\hat{\bar w} - w_k).
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $\hat u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $\hat u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{eq:mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, since $\hat u_k, \hat v_k, C_k > 0$, we have
\begin{equation}
{\rm sign} \frac{\partial \hat{\bar w}}{\partial C_k} = 
{\rm sign} (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

\begin{corollary}
The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
\begin{multline}\label{eq:mfp_K}
%\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
%\end{aligned}
\end{multline}
and
\begin{multline}\label{eq:mfp_sign_threshold}
{\rm sign} \frac{\partial \hat{\bar w}}{\partial \tau} = 
{\rm sign} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
{\rm sign} \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
\end{multline}
\end{corollary}

@@

\begin{corollary}[Reproductive value principle]\label{cor:rvp}
In the notation of Result~\ref{result:mfp},
\begin{equation}
{\rm sign} \frac{\partial \hat{\bar w}}{\partial C_k} = 
{\rm sign} (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

\begin{result}[Evolution of increased genetic variation]
Let $\lambda_1$ be the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $b$ and $B$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
{\rm sign} \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
{\rm sign}(\hat{\bar w} - w_k).
\end{equation}

@@

Now, applying Result~\ref{result:mfp} (Eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
{\rm sign} \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
{\rm sign}\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
\end{equation}
Thus
\begin{equation}
\begin{aligned}
{\rm sign} \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
{\rm sign}\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
{\rm sign}(\hat{\bar w} - w_k),
\end{aligned}
\end{equation}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated. Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural selection.
These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness. 
This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be irreducible. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
We refer to our result as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers that uniformly affect all individuals, the mean fitness here can increase as rates increase.

@@

Since $D$ is a positive diagonal matrix, and by assumption $C_k < 1$ for at least one $k$, $E$ has at least one positive diagonal element, making it aperiodic.  Furthermore, since $D$ is a positive diagonal matrix and $M$ is irreducible, $E$ is also irreducible.  Hence $E$ is primitive.
